MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-20 Manuscript/Manuscript-1-DCA.Rmd
vsn 1

  In this study, we evaluate the precision of Pagel's $\lambda$ to estimate known levels of phylogenetic signal in phenotypic data. First we use computer simulations across differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. For comparison, we also evaluate estimates of $Kappa$ and its empirically-derived effect size ($Z_K$) for this purpose. Additionally, we use simulations to determine how the inclusion of $\lambda$ in phylogenetic regression and ANOVA (i.e., PGLS) affects parameter estimation, and whether levels of phylogenetic signal estimated in a PGLS framework are accurate. We then survey the recent macroevolutionary literature for published papers containing estimates of $\lambda$ from empirical datasets, and compare these empirical estimates to patterns gleaned from our computer simulations. In general we find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. Further, we find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, while effect sizes ($Z_K$) obtained from $Kappa$ display far less variation, and increase in a steady (**uniform?**) manner with increasing phylogenetic signal. We conclude that Pagel's $\lambda$ should not be used for characterizing and comparing the degree of phylogenetic signal in phenotypic datasets, while comparisons of effect sizes obtained from $Kappa$ hold considerable promise in this regard.

@@

Using the simulated phylogenies we estimated the phylogenetic signal in each dataset using Pagel's $\lambda$. Next, for the 1050 datasets at each level of species richness ($n$), we assessed the relationship between $\lambda_{in}$ and $\lambda_{est}$ using regression. Additionally, the precision of $\lambda_{est}$ was approximated by observing both the range and the variation of $\lambda_{est}$ obtained at each input level of phylogenetic signal ($\lambda_{in}$). Finally, for comparison we estimated the degree of phylogenetic signal in each dataset using $Kappa$ [@Blomberg_et_al2003; @Adams2014a], and calculated its effect size [$Z_K$: sensu @AdamsCollyer2016; @AdamsCollyer2019b] for each dataset. Here, the effect size was estimated as: $Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}$, where $K_{obs}$ was the observed $Kappa$, and $\mu_K$ and $\sigma_K$ were the mean and standard deviation of the empirical sampling distribution of values obtained from the permutation distribution [for discussion see: @AdamsCollyer2019]. Because $Kappa$ and $Z_K$ differ in scale from $\lambda$, both measures were linearly normalized to a standard uniform distribution ($0\rightarrow1$), and the range and variation in the set of normalized $Kappa$ and $Z_K$ values at each input level of phylogenetic signal ($\lambda_{in}$) were estimated for comparison. \hfill\break

Our second set of simulations evaluated the extent to which values of $\lambda$ estimated in PGLS regression and ANOVA (i.e., $Y\sim{X}$) corresponded with actual input levels of phylogenetic signal in the response variable. As before we generated 50 pure-birth phylogenies at each of six levels of species richness ($n=2^5 - 2^{10}$), and rescaled each with a set of input $\lambda$ values ($\lambda_{in} = 0.0 - 1.0$; in 21 intervals of 0.05 units). Next, an independent variable $X$ was simulated on each phylogeny under a Brownian motion model of evolution (for PGLS regression). For phylogenetic ANOVA, random groups ($X$) were obtained by simulating a discrete (binary) character on each phylogeny. Next, the dependent variable was simulated in such a manner as to contain a known relationship with $X$ plus random error containing phylogenetic signal. This was accomplished as: $Y=\beta{X}+\epsilon$. Here, the association between $Y$ and $X$  was modeled using a range of values ($\beta=(0,0.25, 0.5, 0.75,1)$), and the residual error was modeled to contain phylogenetic signal simulated under a Brownian motion model of evolution [$\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})$: see @Revell2010 for a similar simulation design]. For each dataset, the fit of the phylogenetic regression was estimated using maximum likelihood, and parameter estimates ($\beta_{est}$ and $\lambda_{est}$) were obtained and evaluated as above. All analyses were performed in R v3.6.0 [@RCT] using the packages `geiger` [@Harmon2008], `caper` [@Orme2013], `phytools` [@Revell2012], and `geomorph` [@AdamsOtarola2013; @AdamsGeomorph]. R-scripts are found in the Supplemental Material.

>>>>

  In this study, we evaluate the precision of Pagel's $\lambda$ to estimate known levels of phylogenetic signal in phenotypic data. First we use computer simulations across differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. For comparison, we also evaluate estimates of $Kappa$ and its empirically-derived effect size ($Z_K$) for this purpose. Additionally, we use simulations to determine how the inclusion of $\lambda$ in phylogenetic regression and ANOVA (i.e., PGLS) affects parameter estimation, and whether levels of phylogenetic signal estimated in a PGLS framework are accurate. We then survey the recent macroevolutionary literature for published papers containing estimates of $\lambda$ from empirical datasets, and compare these empirical estimates to patterns gleaned from our computer simulations. In general we find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. Further, we find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that this variation is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. Alternatively, variation across effect sizes ($Z_K$) obtained from $Kappa$ is far more uniform across the range of input values, and estimates of phylogenetic signal with this measure increase in a uniform manner with increasing phylogenetic signal. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often innacurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure should be treated with caution. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and comparing the strength of phylogenetic signal across datasets.

@@

Using the simulated phylogenies we estimated the phylogenetic signal in each dataset using Pagel's $\lambda$. Next, for the 1050 datasets at each level of species richness ($n$), we assessed the relationship between $\lambda_{in}$ and $\lambda_{est}$ using regression. Additionally, the precision of $\lambda_{est}$ was approximated by observing the variation of $\lambda_{est}$ obtained at each input level of phylogenetic signal ($\lambda_{in}$). Finally, for comparison we estimated the degree of phylogenetic signal in each dataset using $Kappa$ [@Blomberg_et_al2003; @Adams2014a], and calculated its effect size [$Z_K$: sensu @AdamsCollyer2016; @AdamsCollyer2019b] for each dataset. Here, the effect size was estimated as: $Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}$, where $K_{obs}$ was the observed $Kappa$, and $\mu_K$ and $\sigma_K$ were the mean and standard deviation of the empirical sampling distribution of values obtained from the permutation distribution [for discussion see: @AdamsCollyer2019]. Because $Kappa$ and $Z_K$ differ in scale from $\lambda$, both measures were linearly normalized to a standard uniform distribution ($0\rightarrow1$), and the variation in the set of normalized $Kappa$ and $Z_K$ values at each input level of phylogenetic signal ($\lambda_{in}$) were estimated for comparison. \hfill\break

Our second set of simulations evaluated the extent to which values of $\lambda$ estimated in PGLS regression and ANOVA (i.e., $Y\sim{X}$) corresponded with actual input levels of phylogenetic signal in the response variable. As before we generated 50 pure-birth phylogenies at each of six levels of species richness ($n=2^5 - 2^{10}$), and rescaled each with a set of input $\lambda$ values ($\lambda_{in} = 0.0 - 1.0$; in 21 intervals of 0.05 units). Next, an independent variable $X$ was simulated on each phylogeny under a Brownian motion model of evolution (for PGLS regression). For phylogenetic ANOVA, random groups ($X$) were obtained by simulating a discrete (binary) character on each phylogeny. Next, the dependent variable was simulated in such a manner as to contain a known relationship with $X$ plus random error containing phylogenetic signal. This was accomplished as: $Y=\beta{X}+\epsilon$. Here, the association between $Y$ and $X$  was modeled using a range of values: $\beta=(0.0,0.25, 0.5, 0.75,1.0)$, and the residual error was modeled to contain phylogenetic signal simulated under a Brownian motion model of evolution: $\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})$: [see @Revell2010 for a similar simulation design]. For each dataset, the fit of the phylogenetic regression was estimated using maximum likelihood, and parameter estimates ($\beta_{est}$ and $\lambda_{est}$) were obtained and evaluated as above. All analyses were performed in R v3.6.0 [@RCT] using the packages `geiger` [@Harmon2008], `caper` [@Orme2013], `phytools` [@Revell2012], and `geomorph` [@AdamsOtarola2013; @AdamsGeomorph]. R-scripts are found in the Supplemental Material.

### 2020-04-20 Manuscript/Manuscript-1-DCA.tex
vsn 1

In this study, we evaluate the precision of Pagel's \(\lambda\) to
estimate known levels of phylogenetic signal in phenotypic data. First
we use computer simulations across differing numbers of species,
differently shaped phylogenies, and differing input levels of
phylogenetic signal, to explore the degree to which \(\lambda\)
correctly identifies known levels of phylogenetic signal, and under what
circumstances. For comparison, we also evaluate estimates of \(Kappa\)
and its empirically-derived effect size (\(Z_K\)) for this purpose.
Additionally, we use simulations to determine how the inclusion of
\(\lambda\) in phylogenetic regression and ANOVA (i.e., PGLS) affects
parameter estimation, and whether levels of phylogenetic signal
estimated in a PGLS framework are accurate. We then survey the recent
macroevolutionary literature for published papers containing estimates
of \(\lambda\) from empirical datasets, and compare these empirical
estimates to patterns gleaned from our computer simulations. In general
we find that while PGLS parameters (e.g., \(\beta\)) are accurately
estimated with the inclusion of phylogenetic signal, estimates of
\(\lambda\) are not. Further, we find that estimates of \(\lambda\) vary
widely for a given input value of phylogenetic signal, while effect
sizes (\(Z_K\)) obtained from \(Kappa\) display far less variation, and
increase in a steady (\textbf{uniform?}) manner with increasing
phylogenetic signal. We conclude that Pagel's \(\lambda\) should not be
used for characterizing and comparing the degree of phylogenetic signal
in phenotypic datasets, while comparisons of effect sizes obtained from
\(Kappa\) hold considerable promise in this regard.

@@

Using the simulated phylogenies we estimated the phylogenetic signal in
each dataset using Pagel's \(\lambda\). Next, for the 1050 datasets at
each level of species richness (\(n\)), we assessed the relationship
between \(\lambda_{in}\) and \(\lambda_{est}\) using regression.
Additionally, the precision of \(\lambda_{est}\) was approximated by
observing both the range and the variation of \(\lambda_{est}\) obtained
at each input level of phylogenetic signal (\(\lambda_{in}\)). Finally,
for comparison we estimated the degree of phylogenetic signal in each
dataset using \(Kappa\) (Blomberg et al. 2003; Adams 2014a), and
calculated its effect size (\(Z_K\): sensu Adams and Collyer 2016,
2019a) for each dataset. Here, the effect size was estimated as:
\(Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}\), where \(K_{obs}\) was the
observed \(Kappa\), and \(\mu_K\) and \(\sigma_K\) were the mean and
standard deviation of the empirical sampling distribution of values
obtained from the permutation distribution (for discussion see: Adams
and Collyer 2019b). Because \(Kappa\) and \(Z_K\) differ in scale from
\(\lambda\), both measures were linearly normalized to a standard
uniform distribution (\(0\rightarrow1\)), and the range and variation in
the set of normalized \(Kappa\) and \(Z_K\) values at each input level
of phylogenetic signal (\(\lambda_{in}\)) were estimated for comparison.
\hfill\break

Our second set of simulations evaluated the extent to which values of
\(\lambda\) estimated in PGLS regression and ANOVA (i.e., \(Y\sim{X}\))
corresponded with actual input levels of phylogenetic signal in the
response variable. As before we generated 50 pure-birth phylogenies at
each of six levels of species richness (\(n=2^5 - 2^{10}\)), and
rescaled each with a set of input \(\lambda\) values
(\(\lambda_{in} = 0.0 - 1.0\); in 21 intervals of 0.05 units). Next, an
independent variable \(X\) was simulated on each phylogeny under a
Brownian motion model of evolution (for PGLS regression). For
phylogenetic ANOVA, random groups (\(X\)) were obtained by simulating a
discrete (binary) character on each phylogeny. Next, the dependent
variable was simulated in such a manner as to contain a known
relationship with \(X\) plus random error containing phylogenetic
signal. This was accomplished as: \(Y=\beta{X}+\epsilon\). Here, the
association between \(Y\) and \(X\) was modeled using a range of values
(\(\beta=(0,0.25, 0.5, 0.75,1)\)), and the residual error was modeled to
contain phylogenetic signal simulated under a Brownian motion model of
evolution (\(\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})\): see Revell
2010 for a similar simulation design). For each dataset, the fit of the
phylogenetic regression was estimated using maximum likelihood, and
parameter estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were
obtained and evaluated as above. All analyses were performed in R v3.6.0
(R Core Team 2019) using the packages \texttt{geiger} (Harmon et al.
2008), \texttt{caper} (Orme et al. 2013), \texttt{phytools} (Revell
2012), and \texttt{geomorph} (Adams and Otárola-Castillo 2013; Adams et
al. 2020). R-scripts are found in the Supplemental Material.

>>>>

In this study, we evaluate the precision of Pagel's \(\lambda\) to
estimate known levels of phylogenetic signal in phenotypic data. First
we use computer simulations across differing numbers of species,
differently shaped phylogenies, and differing input levels of
phylogenetic signal, to explore the degree to which \(\lambda\)
correctly identifies known levels of phylogenetic signal, and under what
circumstances. For comparison, we also evaluate estimates of \(Kappa\)
and its empirically-derived effect size (\(Z_K\)) for this purpose.
Additionally, we use simulations to determine how the inclusion of
\(\lambda\) in phylogenetic regression and ANOVA (i.e., PGLS) affects
parameter estimation, and whether levels of phylogenetic signal
estimated in a PGLS framework are accurate. We then survey the recent
macroevolutionary literature for published papers containing estimates
of \(\lambda\) from empirical datasets, and compare these empirical
estimates to patterns gleaned from our computer simulations. In general
we find that while PGLS parameters (e.g., \(\beta\)) are accurately
estimated with the inclusion of phylogenetic signal, estimates of
\(\lambda\) are not. Further, we find that estimates of \(\lambda\) vary
widely for a given input value of phylogenetic signal, and that this
variation is not constant across the range of input signal, with
decreased precision when phylogenetic signal is of intermediate
strength. Alternatively, variation across effect sizes (\(Z_K\))
obtained from \(Kappa\) is far more uniform across the range of input
values, and estimates of phylogenetic signal with this measure increase
in a uniform manner with increasing phylogenetic signal. We conclude
that estimates of phylogenetic signal using Pagel's \(\lambda\) are
often innacurate, and thus interpreting strength of phylogenetic signal
in phenotypic datasets based on this measure should be treated with
caution. By contrast, effect sizes obtained from \(Kappa\) hold promise
for characterizing phylogenetic signal, and comparing the strength of
phylogenetic signal across datasets.

@@

Using the simulated phylogenies we estimated the phylogenetic signal in
each dataset using Pagel's \(\lambda\). Next, for the 1050 datasets at
each level of species richness (\(n\)), we assessed the relationship
between \(\lambda_{in}\) and \(\lambda_{est}\) using regression.
Additionally, the precision of \(\lambda_{est}\) was approximated by
observing the variation of \(\lambda_{est}\) obtained at each input
level of phylogenetic signal (\(\lambda_{in}\)). Finally, for comparison
we estimated the degree of phylogenetic signal in each dataset using
\(Kappa\) (Blomberg et al. 2003; Adams 2014a), and calculated its effect
size (\(Z_K\): sensu Adams and Collyer 2016, 2019a) for each dataset.
Here, the effect size was estimated as:
\(Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}\), where \(K_{obs}\) was the
observed \(Kappa\), and \(\mu_K\) and \(\sigma_K\) were the mean and
standard deviation of the empirical sampling distribution of values
obtained from the permutation distribution (for discussion see: Adams
and Collyer 2019b). Because \(Kappa\) and \(Z_K\) differ in scale from
\(\lambda\), both measures were linearly normalized to a standard
uniform distribution (\(0\rightarrow1\)), and the variation in the set
of normalized \(Kappa\) and \(Z_K\) values at each input level of
phylogenetic signal (\(\lambda_{in}\)) were estimated for comparison.
\hfill\break

Our second set of simulations evaluated the extent to which values of
\(\lambda\) estimated in PGLS regression and ANOVA (i.e., \(Y\sim{X}\))
corresponded with actual input levels of phylogenetic signal in the
response variable. As before we generated 50 pure-birth phylogenies at
each of six levels of species richness (\(n=2^5 - 2^{10}\)), and
rescaled each with a set of input \(\lambda\) values
(\(\lambda_{in} = 0.0 - 1.0\); in 21 intervals of 0.05 units). Next, an
independent variable \(X\) was simulated on each phylogeny under a
Brownian motion model of evolution (for PGLS regression). For
phylogenetic ANOVA, random groups (\(X\)) were obtained by simulating a
discrete (binary) character on each phylogeny. Next, the dependent
variable was simulated in such a manner as to contain a known
relationship with \(X\) plus random error containing phylogenetic
signal. This was accomplished as: \(Y=\beta{X}+\epsilon\). Here, the
association between \(Y\) and \(X\) was modeled using a range of values:
\(\beta=(0.0,0.25, 0.5, 0.75,1.0)\), and the residual error was modeled
to contain phylogenetic signal simulated under a Brownian motion model
of evolution: \(\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})\): (see
Revell 2010 for a similar simulation design). For each dataset, the fit
of the phylogenetic regression was estimated using maximum likelihood,
and parameter estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were
obtained and evaluated as above. All analyses were performed in R v3.6.0
(R Core Team 2019) using the packages \texttt{geiger} (Harmon et al.
2008), \texttt{caper} (Orme et al. 2013), \texttt{phytools} (Revell
2012), and \texttt{geomorph} (Adams and Otárola-Castillo 2013; Adams et
al. 2020). R-scripts are found in the Supplemental Material.
